---
title: "Ex 7: Debugging WHAM models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ex 7: Debugging WHAM models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
wham.dir <- find.package("wham")
knitr::opts_knit$set(root.dir = file.path(wham.dir,"extdata"))
library(knitr)
library(kableExtra)
```

In this vignette we show how to use the `do.fit = FALSE` flag to debug a WHAM model.

```{r message=FALSE}
# devtools::install_github("timjmiller/wham", dependencies=TRUE)
library(wham)
```

Load the ASAP3 data file.

```{r}
asap3 <- read_asap3_dat("ex7_SNEMAYT.dat")
```

Prepare the WHAM model (`m3` from [example 1](https://timjmiller.github.io/wham/articles/ex1_SNEMA_yellowtail_flounder.html)).

```{r}
input <- prepare_wham_input(asap3, recruit_model=2, model_name="Ex 1: SNEMA Yellowtail Flounder",
                              NAA_re = list(sigma="rec+1", cor="iid"))
```

Try to fit the model... uh oh.

```{r error=TRUE}
mod <- fit_wham(input, do.osa = F, do.retro = F)
```

What's wrong? It looks like the likelihood function is returning `NaN`. It is often easier to diagnose problems like this using the *unoptimized* model, i.e. look at everything using the initial parameter values. WHAM includes a `do.fit = F` flag in `fit_wham` to return the unoptimized model object returned by `TMB::MakeADFun`. Let's see how it works.

```{r}
mod <- fit_wham(input, do.fit = F)
```

This runs without an error. The optimization failed because the likelihood was `NaN`, and now we can see which of the likelihood components was responsible. To do so, we need to look at the `REPORT`ed objects with `"nll"` in their name. We can use the `$report()` function from TMB.

```{r}
therep = mod$report()
```

See all of the `REPORT`ed objects.

```{r}
names(therep)
```

Now just get the objects with `"nll"` in their name, and sum over all individual values.

```{r}
sapply(grep("nll",names(therep),value=T), function(x) sum(therep[[x]]))
```

The likelihood components that are equal to 0 are not used in the model (no random effects on `M`, `Ecov`, `selectivity`, etc.). `nll` is the total likelihood and is `NaN`. The one troublesome component is `nll_agg_catch`. This is the total (aggregate) catch from the fishery in each year. It could be an issue with the catch data or the model-predicted catch, since both are in the likelihood calculation.

Search the [WHAM .cpp file](https://github.com/timjmiller/wham/blob/master/src/wham_v0.cpp) for "nll_agg_catch". We find that on line 888, this depends on `agg_catch` (the catch data) and `pred_catch` (model-predicted catch).

The catch data looks ok.

```{r}
input$data$agg_catch
```

The predicted catch is calculated on lines 879-880. `pred_catch` depends on `NAA` (numbers-at-age), `FAA` (F at age), `ZAA` (Z at age), and `waa` (weight-at-age data).

`NAA` looks ok.
```{r}
therep$NAA
```

`FAA` looks ok - no issues with `F` or selectivity.
```{r}
therep$FAA[,1,] # middle dim is n.fleets = 1
```

`ZAA` looks ok - no issues with `M` either.
```{r}
therep$ZAA
```

Ah, here is the problem. The weight at age data has an entry of `-99`. This means that `pred_catch` is negative on line 880, and we take the log of a negative number on line 883.

```{r}
input$data$waa[1,,]
```

If we fix this issue with the data file, the model runs fine!
